Review Paper Draft

### 2020-08-15 main.tex
All sections: updating citation and figure refs.

As with multiple linear regression models, Lasso penalized regression assumes independent observations and uncorrelated errors. In the presence of population structure or cryptic relatedness, however, samples are correlated, violating this assumption. If unaccounted for, this dependency can result in the identification of spurious associations. Though methods to correct for population stratification and cryptic relatedness have been extensively investigated \citep{Amin2007, hoffman2013correcting, price2006principal, Rakitsch2012, bhatnagar2019simultaneous, Sillanpaeae2011}, until recently the majority of this work has focused on plant and animal breeding data, much of which is experimentally controlled. While many standard correction techniques may in principal be applied to human genetic data, their performance has not been well-studied within the context of human ancestry and non-experimental conditions \citep{lawson2019population, barton2019population}.

One reason that population stratification warrants such concern in genetic studies is that due to ethnic and geographic segregation, population stratification tends to be associated with differing environmental exposures and cultural practices.  In other words, random genetic variation is likely to be correlated with non-genetic factors that also influence the trait of interest, thereby confounding the genotype-phenotype relationship and leading to spurious associations and biased estimates of SNP effects. Biased estimates of this nature severely hinder prediction and understanding differences among populations \citep{barton2019population}. Although bias at individual loci may be small, this bias becomes magnified when aggregated across thousands of SNPs, as is done when calculating polygenic risk scores \citep{barton2019population}.

In order to clarify and emphasize the importance of the subtle distinction between population structure and environmental heterogeneity, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations \citep{Sillanpaeae2011, sul2018population}, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted lasso penalized regression (PC-Lasso) and lasso penalized multivariate LMMs (LMM-Lasso). We review the statistical details of these methods as well as the consequences \pb{do we review \emph{consequences} in section 3?} \anna{tbd} of the ways in which they attempt to correct for population structure and environmental heterogeneity. In section \ref{sec:results}, we illustrate the concepts described in section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.

@@

Population structure is a common phenomenon in genetic data. Varying levels of relatedness are almost always present among genetic samples, even in samples of unrelated individuals and seemingly homogeneous populations. For example, European American \citep{campbell2005demonstrating}, Han Chinese \citep{xu2009genomic, chen2009genetic}, and most recently, cohorts within the UK Biobank data \citep{haworth2019apparent} have been shown to exhibit patterns of population and geographic structure despiet their seemingly similar subjects.

@@

Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. However, in this review we focus on methods to account for unobserved relatedness. Throughout this paper, the terms \textbf{population structure} and \textbf{structured population} will be used to describe a sample of individuals in which population stratification, cryptic relatedness, or both are present. Population stratification in particular has been of great concern in genetic studies due to its potential to lead to spurious associations. This phenomenon is commonly described as \textbf{confounding due to population stratification or ancestry}. The mechanism of this phenomenon warrants some discussion as it is often overlooked that population structure itself does not confound the genotype-phenotype relationship in the classical sense. While population structure may directly affect observed allele frequencies, there must also exist some mechanism by which population stratification affects the phenotype \citep{barton2019population, vilhjalmsson2012nature}, namely, the environment.

@@

This mechanism of environmental confounding is summarized in the directed acyclic graph displayed in figure \ref{fig:ps_env} \pb{typically, this is capitalized so it matches how Figure 1 appears in the caption}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors directly influence population-specific allele frequencies, or vice versa, only that the two may be related. In the context of our air quality example, this means that we assume air quality is not directly altering individuals' allele frequencies, nor are allele frequencies causing individuals to move to regions of better or worse air quality. Genetic relatedness and environmental effects may become associated over time for any number of complex reasons beyond the scope of this paper. However, as we will discuss, many methods of correcting for population structure rely on this relationship and do not formally differentiate between effects of population structure and those of environment. Such methods assume that genetic relatedness serves as an accurate proxy for environmental exposures, which may or may not be the case depending on the exposure and population in question. \pb{is this an important point?} \anna{hmm... maybe? robust to this assumption: if it's not a good proxy it's not going to cause confounding}

@@

Statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. In order to reduce confounding these methods attempt to block the causal pathway between population structure and observed SNP data. This is illustrated by the blocked arrow in figure \ref{fig:ps_env_block}. These methods do not directly address the environmental component of confounding due to population structure. Nevertheless, they may be effective if they are able to indirectly block the causal pathway between the environment and observed SNPs. Whether or not they are effective depends upon whether population structure can be accurately inferred, and if this is a good proxy for environmental effects. \anna{maybe say something about how 'if it's a good proxy - great! they'll work', but 'if it's not a good proxy, also fine! because then it's not a confounder! ...SE's potentially a problem like in GLS... but this is different with lasso. (maybe not worth getting into here)}

@@

Multivariate mixed model approaches reduce these problems by modeling the relationship between a phenotype of interest and all SNPs simultaneously via penalized regression, while controlling for dependent error structures  \citep{Rakitsch2012, bhatnagar2019simultaneous}. This approach assumes there are a relatively small number of SNPs associated with the phenotype of interest that have large effect sizes. These SNPs will be included in the model as both fixed and random effects, while SNPs with smaller effect sizes are modeled only as part of the random effect in order to control for dependent errors due to relatedness and sample structure.

The remainder of this review will focus on the performance of such multivariate Lasso-penalized methods for the analysis of continuous, normally distributed outcomes. Specifically, PC-adjustment and LMM methods, which we will refer to as PC-Lasso and LMM-Lasso, respectively. We consider two similar implementations of the LMM-Lasso method, one developed by \citep{Rakitsch2012} and a more recent implementation from \citep{bhatnagar2019simultaneous}. Subsequent references to LMM-Lasso apply to both implementations, unless otherwise stated. Where necessary, we will differentiate between these implementations by referring to them as LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, respectively.

%In section \ref{sec:results}, we compare the performance of PC-Lasso and LMM-Lasso in terms of their abilities to accurately estimate SNP effects and in terms of their true sign rates in the presence of varying levels of relatedness and environmental confounding. 
Although several studies have shown that LMMs outperform PC-adjustment methods in the univariate framework \citep{wang2013analytical, kang2010variance, zhao2007arabidopsis}, to the best of our knowledge this is the only comprehensive review and comparison of these methods in a penalized multivariate framework and using a simulation scheme motivated by human genetic relatedness and the environmental mechanism of confounding due to population structure.\\

@@

To describe and compare PC-Lasso and LMM-Lasso, we consider an $n \times p$ matrix of genotype data $\boldsymbol{G}$ for $n$ subjects and $p$ SNPs, where each $G_{i,j} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$. We denote the normalized version of this genotype matrix as $\bX$, where each $X_{ij} = (G_{ij} - 2 p_j) / \sqrt{p_j (1 - p_j)}$ and $p_j$ is the minor allele frequency (MAF) of SNP $j$ \citep{zhang2015principal, price2006principal}. Additionally, $\bX_0$ represents an $n \times p_0$ matrix of unpenalized covariates and the intercept. Thus, at a minimum, $\bX_0$ will be an $n \times 1$ vector of 1s to represent the intercept, in which case $\bbeta_0 \equiv \beta_0$ is the scalar intercept. However, $\bX_0$ may additionally include covariates such as age and sex, which are of interest to include in the model without penalty. We assume that the $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of these covariates, additive SNP effects, environmental effects, and observational noise, and consider the linear mixed model,
\begin{equation}
    \label{eqn:model}
    \by = \bX_0 \bbeta_0 + \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\bbeta_0$ is a $p_0 \times 1$ vector which includes the intercept term and the effects of any unpenalized covariates, $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bgamma$ is a $q \times 1$ vector of environmental effects, $\bZ$ is an $n \times q$ matrix that allocates environmental effects among subjects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$. Often it is assumed that subjects belong to $q \le p$ subpopulations, and $\bZ$ is a corresponding incidence matrix of 0s and 1s. However, as we will discuss, this is not the only possibility.

@@

We have so far used the term $\bW$ to describe the LMM-Lasso method. However, the form of $\bW$ and method of variance component estimation differs between LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. LMM-Lasso-Rakitsch \citep{Rakitsch2012} uses a two-step procedure for estimating the variance components, $\sigmaee$ and $\delta := \sigmaee / \sigmagg$. These variance components and the corresponding weights are estimated once under the assumption of a null model. A Lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \citep{glmnet} or \texttt{ncvreg} \citep{ncvreg}. LMM-Lasso-ggmix \citep{bhatnagar2019simultaneous} proposes iteratively updating $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm. This iterative update scheme necessitates the use of the corresponding \texttt{R} package \texttt{ggmix} \citep{ggmix, bhatnagar2019simultaneous}. We postpone further discussion of the differences between these methods to section \ref{sec:results}.

@@

It has been shown that the random effect formulation of model \eqref{eqn:model} used by LMM-Lasso can be equivalently expressed in terms of fixed effects. \citep{zhang2015principal} derived this equivalence using a probabilistic PCA formulation, while \citep{hoffman2013correcting} used the singular value decomposition of $\bX$, and the decomposition of $\bK$ \anna{which we briefly review here...}

@@

In carrying out simulation studies of the LMM regression model \eqref{eqn:lmm_reg}, a common practice \citep{Rakitsch2012, bhatnagar2019simultaneous} is to draw a new value of the quantity $\bX\bgamma$ from a $N(\boldsymbol{0}, \sigmagg \bK)$ for each replication of the analysis.  We argue that this is somewhat misleading from the perspective of understanding the statistical properties of the method. Averaging over situations in which the environmental confounding is constantly changing direction and magnitude is inconsistent with the confounding environmental effects. This has the effect of ``washing out'' bias in $\bbetaHat$, which is of interest, while inflating the variance. To illustrate this concept, consider the low-dimensional ordinary least squares (OLS) estimates of $\bbeta$ based on model \eqref{eqn:lmm_reg} where, without loss of generatlity, we assume $\bX_0 \bbeta_0 = \boldsymbol{0}$ for ease of notation: 
\begin{align}
    \mathbb{E}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\by) \notag \\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\bX\bbeta + \bX\bgamma + \beps) \label{eqn:expectation}\\
    &\quad \notag\\
    \mathbb{V}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\by) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bbeta + \bX\bgamma + \beps) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bgamma + \beps) \bX  (\bX^\T \bX)^{-1} \label{eqn:variance}.
\end{align}
Under the assumption that environmental effects arise from a random process, $\bgamma \sim N(\boldsymbol{0}, p^{-1} \sigmagg I_p)$, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta$, and $\mathbb{V}(\bbetaHat) = p^{-1}\sigmagg\bI + \sigmaee (\bX^\T \bX)^{-1}$, respectively. Consider our earlier example involving confounding due to air quality. Simulating $\bgamma$ as a random quantity for each replication corresponds to subpopulation A being exposed to good air quality in one replication, and bad air quality in the next. Reporting the average across these situations is inconsistent with the question we are truly interested in, namely: what would happen if we took repeated samples from populations A and B, when these populations are subject to different environmental exposures? Those repeated samples would show a systematic bias due to the difference in air quality, but this is masked if $\bgamma$ is assumed to be random. Treating $\bgamma$ as a fixed quantity so that environmental effects are consistent across replications more closely approximates environmental confounding. Under the assumption that the quantity $\bgamma$ is fixed, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta + \bgamma, \mathbb{V}(\bbetaHat) = \sigmaee (\bX^\T \bX)^{-1}$, allowing us to quantify estimation bias due to environmental heterogeneity.

@@

@@

Given the nature of environmental confounding in our setting of interest, and in order to assess these methodological performance in terms of estimation accuracy and bias specifically, we simulate according to a fixed $\bgamma$ framework using the following data-generating model
\begin{equation}
    \label{eqn:sim}
    \by = a \bX \bbeta + b \bZ \bgamma + c \beps,
\end{equation}
where $\by$ represents the pseudophenotype, and $a, b, c$ are scaling factors determined such that the variance of $\by$ is equal to 1 and is partitioned into genetic signal, environmental confounding, and random noise components,
\begin{equation}
    \label{eqn:parition_y}
    \mathbb{V}(\by) = \eta + (1 - \eta) \xi + (1-\eta)(1-\xi).
\end{equation}
$\eta$ defines the proportion of the variance of $\by$ attributable to additive genetic factors, commonly referred to as the narrow-sense heritability, $h^2 = \sigmagg/(\sigmagg + \sigmaee)$. $\eta$ also describes the data-generating model's signal to noise ratio (SNR). $\xi$ partitions the remaining variance of $\by$ into latent environmental heterogeneity and random noise. So, for example $\eta = 0.5$ and $\xi = 0.8$ corresponds to a 1:1 SNR, and means that 50\% of the variability in $\by$ is due to additive genetic effects, and 40\% due to environmental effects. \anna{more details about how/why this lambda selection method - cite CV as bad for LMMs, BIC not great bc different likelihoods} The $\lambda$ that allowed for the first 50 variables to enter the model was selected for each method. 1000 simulations were performed for each unique combination of parameters, datatype, subpopulation structure, and environmental effect structure.

@@

\subsubsection{Design matrix generation}
We generated these pseudophenotypes using three types of high-dimensional ($p > n$) $\bX$ matrices: (1) simulated SNP data from independent subpopulations, (2) simulated SNP data from admixed subpopulations, and (3) real genome-wide SNP data from known caucasian ($n = 417$), African American ($n = 134$), and Hispanic ($n = 37$) subpopulations. \anna{cite Vanderwas data} $p = 1000$, $n = \{200, 400, 600, 800 \}$ for the simulated data types, and 50 causal SNPs were used throughout.

\subsubsection{Subpopulation structure}
For the fully simulated data, two subpopulation structures were considered: (1) fine and (2) coarse. We define a coarse structure to be one characterized by a few large subpopulations intended to model large-scale environmental heterogeneity and population stratification, such as that described in our motivating air quaility example. Conversely, we define a fine subpopulation structure to be one characterized by many small subpopulations, or founding populations, which corresponds to a finer grained continuum of environmental heterogeneity. Whether fine or coarse, the simulated data were generated such that subpopulations, or founding populations, are of equal size. We note that the fine and coarse subpopulation structures are distinct from the level of admixture in the simulated data. For example, the discrete nonoverlapping subpopulations in the simulated independent subpopulation data result in an RRM which more directly corresponds to the discrete and fixed environmental effects, fine or coarse, than do the continuous and overlapping subpopulations in the simulated admixture data. The empirical data is comprised of three racial subpopulations; one relatively large, one small, and one in the middle. Thus it represents a hybrid of our simulated coarse and fine structures, with little apparent admixture. Supplementary figures \anna{?, ?, and ?} illustrate the simulated kinship structures in the fine and coarse subpopulation settings, as well as that present in the empirical SNP data.

@@

 Our initial hypothesis was that structures of $\bgamma$ that can be well-approximated by a linear combination of the observed genetic data would be well-approximated and corrected for by PC-Lasso, whereas LMM-Lasso may be more able to correct for subpopulation-specific effects that are nonlinear in nature and as such do not lend themselves to such an approximation. To that end we investigated two forms of environmental effect magnitudes: (1) exponential, and (2) dichotomous. In addition to the magnitudes of effects, we also considered whether individuals in genetically similar subpopulations would experience similar environmental effects. For example, a geographic environmental exposure such as the air quality described in our motivating example is likely to be similar among genetically related individuals due to concordance between genetic clusters and geography. However, behavior-based environmental exposures may be less likely to be shared by subpopulations which are genetically and geographically similar. As an example, \citep{ng2014smoking} found considerable within-region variation in smoking rates in Africa, Asia, and South America.

>>>>

As with multiple linear regression models, Lasso penalized regression assumes independent observations and uncorrelated errors. In the presence of population structure or cryptic relatedness, however, samples are correlated, violating this assumption. If unaccounted for, this dependency can result in the identification of spurious associations. Though methods to correct for population stratification and cryptic relatedness have been extensively investigated, until recently the majority of this work has focused on plant and animal breeding data, much of which is experimentally controlled \citep{Amin2007, hoffman2013correcting, price2006principal, Rakitsch2012, bhatnagar2019simultaneous, Sillanpaeae2011}. While many standard correction techniques may in principal be applied to human genetic data, their performance has not been well-studied within the context of human ancestry and non-experimental conditions \citep{lawson2019population, barton2019population}.

One reason that population stratification warrants such concern in genetic studies is that due to ethnic and geographic segregation, population stratification tends to be associated with differing environmental exposures and cultural practices \citep{thornton2015statistical}.  In other words, random genetic variation is likely to be correlated with non-genetic factors that also influence the trait of interest, thereby confounding the genotype-phenotype relationship and leading to spurious associations and biased estimates of SNP effects. Biased estimates of this nature severely hinder prediction and understanding differences among populations \citep{barton2019population}. Although bias at individual loci may be small, this bias becomes magnified when aggregated across thousands of SNPs, as is done when calculating polygenic risk scores \citep{barton2019population, peterson2019genome}.

In order to clarify and emphasize the importance of the subtle distinction between population structure and environmental heterogeneity, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures \citep{Sillanpaeae2011, sul2018population, vilhjalmsson2012nature, barton2019population}. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted lasso penalized regression (PC-Lasso) and lasso penalized multivariate LMMs (LMM-Lasso). We review the statistical details of these methods as well as the consequences \pb{do we review \emph{consequences} in section 3?} \anna{tbd} of the ways in which they attempt to correct for population structure and environmental heterogeneity. In section \ref{sec:results}, we illustrate the concepts described in section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.

@@

Population structure is a common phenomenon in genetic data. Varying levels of relatedness are almost always present among genetic samples, even in samples of unrelated individuals and seemingly homogeneous populations. For example, European American, Han Chinese, and most recently, cohorts within the UK Biobank data have been shown to exhibit patterns of population and geographic structure despiet their seemingly similar subjects \citep{campbell2005demonstrating, xu2009genomic, chen2009genetic, haworth2019apparent}.

@@

Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. However, in this review we focus on methods to account for unobserved relatedness. Throughout this paper, the terms \textbf{population structure} and \textbf{structured population} will be used to describe a sample of individuals in which population stratification, cryptic relatedness, or both are present. Population stratification in particular has been of great concern in genetic studies due to its potential to lead to spurious associations. This phenomenon is commonly described as \textbf{confounding due to population stratification or ancestry}. The mechanism of this phenomenon warrants some discussion as it is often overlooked that population structure itself does not confound the genotype-phenotype relationship in the classical sense. While population structure may directly affect observed allele frequencies, there must also exist some mechanism by which population stratification affects the phenotype, namely, the environment \citep{barton2019population, vilhjalmsson2012nature}.

@@

This mechanism of environmental confounding is summarized in the directed acyclic graph displayed in Figure \ref{fig:ps_env}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors directly influence population-specific allele frequencies, or vice versa, only that the two may be related. In the context of our air quality example, this means that we assume air quality is not directly altering individuals' allele frequencies, nor are allele frequencies causing individuals to move to regions of better or worse air quality. Genetic relatedness and environmental effects may become associated over time for any number of complex reasons beyond the scope of this paper. However, as we will discuss, many methods of correcting for population structure rely on this relationship and do not formally differentiate between effects of population structure and those of environment. Such methods assume that genetic relatedness serves as an accurate proxy for environmental exposures, which may or may not be the case depending on the exposure and population in question. \pb{is this an important point?} \anna{hmm... maybe? robust to this assumption: if it's not a good proxy it's not going to cause confounding}

@@

Statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. In order to reduce confounding these methods attempt to block the causal pathway between population structure and observed SNP data. This is illustrated by the blocked arrow in Figure \ref{fig:ps_env_block}. These methods do not directly address the environmental component of confounding due to population structure. Nevertheless, they may be effective if they are able to indirectly block the causal pathway between the environment and observed SNPs. Whether or not they are effective depends upon whether population structure can be accurately inferred, and if this is a good proxy for environmental effects. \anna{maybe say something about how 'if it's a good proxy - great! they'll work', but 'if it's not a good proxy, also fine! because then it's not a confounder! ...SE's potentially a problem like in GLS... but this is different with lasso. (maybe not worth getting into here)}

@@

Multivariate mixed model approaches reduce these problems by modeling the relationship between a phenotype of interest and all SNPs simultaneously via penalized regression, while correcting for dependent error structures  \citep{Rakitsch2012, bhatnagar2019simultaneous}. This approach assumes there are a relatively small number of SNPs associated with the phenotype of interest that have large effect sizes. These SNPs will be included in the model as both fixed and random effects, while SNPs with smaller effect sizes are modeled only as part of the random effect in order to control for dependent errors due to relatedness and sample structure.

The remainder of this review will focus on the performance of such multivariate Lasso-penalized methods for the analysis of continuous, normally distributed outcomes. Specifically, PC-adjustment and LMM methods, which we will refer to as PC-Lasso and LMM-Lasso, respectively. We consider two similar implementations of the LMM-Lasso method, one developed by \citet{Rakitsch2012} and a more recent implementation from \citet{bhatnagar2019simultaneous}. Subsequent references to LMM-Lasso apply to both implementations, unless otherwise stated. Where necessary, we will differentiate between these implementations by referring to them as LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, respectively.

Several studies have shown that LMMs outperform PC-adjustment methods in the univariate framework \citep{wang2013analytical, kang2010variance, zhao2007arabidopsis}. However, to the best of our knowledge this is the only comprehensive review and comparison of these methods in a penalized multivariate framework and using a simulation scheme motivated by human genetic relatedness and the environmental mechanism of confounding due to population structure.\\

@@

To describe and compare PC-Lasso and LMM-Lasso, we consider an $n \times p$ matrix of genotype data $\boldsymbol{G}$ for $n$ subjects and $p$ SNPs, where each $G_{i,j} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$. We denote the standardized version of this genotype matrix as $\bX$, where each $X_{ij} = (G_{ij} - 2 p_j) / \sqrt{2 p_j (1 - p_j)}$ and $p_j$ is the minor allele frequency (MAF) of SNP $j$ \citep{zhang2015principal, price2006principal}. Additionally, $\bX_0$ represents an $n \times p_0$ matrix of unpenalized covariates and the intercept. Thus, at a minimum, $\bX_0$ will be an $n \times 1$ vector of 1s to represent the intercept, in which case $\bbeta_0 \equiv \beta_0$ is the scalar intercept. However, $\bX_0$ may additionally include covariates such as age and sex, which are of interest to include in the model without penalty. We assume that the $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of these covariates, additive SNP effects, environmental effects, and observational noise, and consider the linear mixed model,
\begin{equation}
    \label{eqn:model}
    \by = \bX_0 \bbeta_0 + \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\bbeta_0$ is a $p_0 \times 1$ vector which includes the intercept term and the effects of any unpenalized covariates, $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bgamma$ is a $q \times 1$ vector of environmental effects, $\bZ$ is an $n \times q$ matrix that allocates environmental effects among subjects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$. Often it is assumed that subjects belong to $q \le p$ subpopulations, and $\bZ$ is a corresponding incidence matrix of 0s and 1s. However, as we will discuss, this is not the only possibility.

@@

We have so far used the term $\bW$ to describe the LMM-Lasso method. However, the form of $\bW$ and method of variance component estimation differs between LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. LMM-Lasso-Rakitsch uses a two-step procedure for estimating the variance components, $\sigmaee$ and $\delta := \sigmaee / \sigmagg$ \citep{Rakitsch2012}. These variance components and the corresponding weights are estimated once under the assumption of a null model. A Lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \citep{glmnet} or \texttt{ncvreg} \citep{ncvreg}. LMM-Lasso-ggmix iteratively updates $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm \citep{bhatnagar2019simultaneous}. This iterative update scheme necessitates the use of the corresponding \texttt{R} package \texttt{ggmix} \citep{ggmix}. We postpone further discussion of the differences between these methods to section \ref{sec:results}.

@@

It has been shown that the random effect formulation of model \eqref{eqn:model} used by LMM-Lasso can be equivalently expressed in terms of fixed effects. \citet{zhang2015principal} derived this equivalence using a probabilistic PCA formulation, while \citet{hoffman2013correcting} used the singular value decomposition of $\bX$, and the decomposition of $\bK$ \anna{which we briefly review here...}

@@

In carrying out simulation studies of the LMM regression model \eqref{eqn:lmm_reg}, a common practice is to draw a new value of the quantity $\bX\bgamma$ from a $N(\boldsymbol{0}, \sigmagg \bK)$ for each replication of the analysis \citep{Rakitsch2012, bhatnagar2019simultaneous}.  We argue that this is somewhat misleading from the perspective of understanding the statistical properties of the method. Averaging over situations in which the environmental confounding is constantly changing direction and magnitude is inconsistent with the confounding environmental effects. This has the effect of ``washing out'' bias in $\bbetaHat$, which is of interest, while inflating the variance. To illustrate this concept, consider the low-dimensional ordinary least squares (OLS) estimates of $\bbeta$ based on model \eqref{eqn:lmm_reg} where, without loss of generatlity, we assume $\bX_0 \bbeta_0 = \boldsymbol{0}$ for ease of notation: 
\begin{align}
    \mathbb{E}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\by) \notag \\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\bX\bbeta + \bX\bgamma + \beps) \label{eqn:expectation}\\
    &\quad \notag\\
    \mathbb{V}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\by) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bbeta + \bX\bgamma + \beps) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bgamma + \beps) \bX  (\bX^\T \bX)^{-1} \label{eqn:variance}.
\end{align}
Under the assumption that environmental effects arise from a random process, $\bgamma \sim N(\boldsymbol{0}, p^{-1} \sigmagg I_p)$, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta$, and $\mathbb{V}(\bbetaHat) = p^{-1}\sigmagg\bI + \sigmaee (\bX^\T \bX)^{-1}$, respectively. Consider our earlier example involving confounding due to air quality. Simulating $\bgamma$ as a random quantity for each replication corresponds to subpopulation A being exposed to good air quality in one replication, and bad air quality in the next. Reporting the average across these situations is inconsistent with the question we are truly interested in, namely: what would happen if we took repeated samples from populations A and B, when these populations are subject to different environmental exposures? Those repeated samples would show a systematic bias due to the difference in air quality, but this is masked if $\bgamma$ is assumed to be random. Treating $\bgamma$ as a fixed quantity so that environmental effects are consistent across replications more closely approximates environmental confounding. Under the assumption that the quantity $\bgamma$ is fixed, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta + \bgamma, \mathbb{V}(\bbetaHat) = \sigmaee (\bX^\T \bX)^{-1}$, allowing us to quantify estimation bias due to environmental heterogeneity.

@@

\anna{bring this back to why it may not matter if genetic similarity isn't a good proxy for population structure - in that case it's not a confounder anyway}

@@

Given the nature of environmental confounding in our setting of interest, and in order to assess these methodological performance in terms of estimation accuracy and bias specifically, we simulate according to a fixed $\bgamma$ framework using the following data-generating model
\begin{equation}
    \label{eqn:sim}
    \by = a \bX \bbeta + b \bZ \bgamma + c \beps,
\end{equation}
where $\by$ represents the pseudophenotype, and $a, b, c$ are scaling factors determined such that the variance of $\by$ is equal to 1 and is partitioned into genetic signal, environmental confounding, and random noise components,
\begin{equation}
    \label{eqn:parition_y}
    \mathbb{V}(\by) = \eta + (1 - \eta) \xi + (1-\eta)(1-\xi).
\end{equation}
$\eta$ defines the proportion of the variance of $\by$ attributable to additive genetic factors, commonly referred to as the narrow-sense heritability, $h^2 = \sigmagg/(\sigmagg + \sigmaee)$. $\eta$ also describes the data-generating model's signal to noise ratio (SNR). $\xi$ partitions the remaining variance of $\by$ into latent environmental heterogeneity and random noise. So, for example $\eta = 0.5$ and $\xi = 0.8$ corresponds to a 1:1 SNR, and means that 50\% of the variability in $\by$ is due to additive genetic effects, and 40\% due to environmental effects. \anna{more details about how/why this lambda selection method - cite CV as bad for LMMs, BIC not great bc different likelihoods... tie this into evaluating performance of each method in terms of variable selection} The $\lambda$ that allowed for the first 50 variables to enter the model was selected for each method. 1000 simulations were performed for each unique combination of parameters, datatype, subpopulation structure, and environmental effect structure.

@@

\subsubsection{Design matrix generation}
We generated these pseudophenotypes using three types of high-dimensional ($p > n$) $\bX$ matrices: (1) simulated SNP data from independent subpopulations, (2) simulated SNP data from admixed subpopulations, and (3) real genome-wide SNP data from known caucasian ($n = 417$), African American ($n = 134$), and Hispanic ($n = 37$) subpopulations \anna{cite Vanderwas data}. $p = 1000$, $n = \{200, 400, 600, 800 \}$ for the simulated data types, and 50 causal SNPs were used throughout.

\subsubsection{Subpopulation structure}
For the fully simulated data, two subpopulation structures were considered: (1) fine and (2) coarse. We define a coarse structure to be one characterized by a few large subpopulations intended to model large-scale environmental heterogeneity and population stratification, such as that described in our motivating air quaility example. Conversely, we define a fine subpopulation structure to be one characterized by many small subpopulations, or founding populations, which corresponds to a finer grained continuum of environmental heterogeneity. Whether fine or coarse, the simulated data were generated such that subpopulations, or founding populations, are of equal size. We note that the fine and coarse subpopulation structures are distinct from the level of admixture in the simulated data. For example, the discrete nonoverlapping subpopulations in the simulated independent subpopulation data result in an RRM which more directly corresponds to the discrete and fixed environmental effects, fine or coarse, than do the continuous and overlapping subpopulations in the simulated admixture data. The empirical data is comprised of three racial subpopulations; one relatively large, one small, and one in the middle. Thus it represents a hybrid of our simulated coarse and fine structures, with little apparent admixture. Supplementary Figures \anna{?, ?, and ?} illustrate the simulated kinship structures in the fine and coarse subpopulation settings, as well as that present in the empirical SNP data.

@@

 Our initial hypothesis was that structures of $\bgamma$ that can be well-approximated by a linear combination of the observed genetic data would be well-approximated and corrected for by PC-Lasso, whereas LMM-Lasso may be more able to correct for subpopulation-specific effects that are nonlinear in nature and as such do not lend themselves to such an approximation. To that end we investigated two forms of environmental effect magnitudes: (1) exponential, and (2) dichotomous. In addition to the magnitudes of effects, we also considered whether individuals in genetically similar subpopulations would experience similar environmental effects. For example, a geographic environmental exposure such as the air quality described in our motivating example is likely to be similar among genetically related individuals due to concordance between genetic clusters and geography. However, behavior-based environmental exposures may be less likely to be shared by subpopulations which are genetically and geographically similar. As an example, \citet{ng2014smoking} found considerable within-region variation in smoking rates in Africa, Asia, and South America.
